Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-16 main.tex
Compress gaussian variance and add gaussian
Add bound for Gamma(rate) distribution

@@

\subsection{Full Gaussian}
To model $\cN(m,\sigma^2)$, we take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
The log-density is written
\begin{align}
	\log p(X) 
	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
\end{align}
From this formula, we can identify the duality map
\begin{align}
	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
\end{align}
and we can also explicit the log-partition as a function of the natural parameters
\begin{align}
	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
\end{align}
We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
\begin{align}
	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
\end{align}
Now let us recover the entropy
\begin{align}
	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
	& = - \half \log (\mu_2 - \mu_1^2) + \cst \; .
\end{align}
In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.
Note that this entropy is not self-concordant. Taking derivatives in the direction of $\mu_1$, and setting $\mu = (4, 16)$,  the self-concordance inequality is not verified.

@@

\textbf{When $n\geq 2$}, we get 
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + n}{a +n - 2} -1
	= \frac{n_0 - a + 2}{a +n - 2}
	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
\end{align}
so putting it all together we get 
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm)} 
	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \frac{n_0 - a + 2}{(a+n-2)(n_0+n)} \\
	&= \frac{1}{a+n-2} + \frac{(n_0 - a)^2 + 2(n_0-a) \pm 1}{2(a+n-2)(n_0+n)} \\
	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
\end{align}

\textbf{When $n=1$} the bound~\eqref{eq:lemma_natp_bound}  on the expected natural parameter gives
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + 1}{a} -1
	= \frac{n_0 +1 - a}{a}
\end{align}
so putting it all together we get
\begin{align}
	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 - a + 1}{a} \\
	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
\end{align}
so we do recover the same bias variance structure as when $n\geq2$.
\end{proof}

>>>>

\section{Gamma with known shape}

We are looking for $\Gamma(\alpha, \beta^*)$, where the shape $\alpha$ is known, but the rate $\beta^*$ is unknown. The logpartition is $\logpart(\beta) = - \alpha \log(\beta)$ and the mean parameter is $\mu* = \frac{\alpha}{\beta}$.
The symmetrized Bregman of the MAP verifies the following upper bound
\alignn{
    \forall n\geq \inv{\alpha},\quad
	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
	\leq\frac{1}{n_0+n} + 
	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - 1 - \inv{\alpha n_0}}^2}{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} + \frac{n - \inv{\alpha}}{n_0}}}
}
where the bias vanish when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.

@@

\subsection{Full Gaussian}
To model $\cN(m,\sigma^2)$, we take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
The log-density is written
\begin{align}
	\log p(X) 
	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
\end{align}
From this formula, we can identify the duality map
\begin{align}
	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
\end{align}
and we can also explicit the log-partition as a function of the natural parameters
\begin{align}
	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-2\natp_2) \; .
\end{align}
We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
\begin{align}
	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
\end{align}
Now let us recover the entropy
\begin{align}
	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
	& = - \half \log (\mu_2 - \mu_1^2) + \cst \; .
\end{align}
In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.
Note that this entropy is not self-concordant. Taking derivatives in the direction of $\mu_1$, and setting $\mu = (4, 16)$,  the self-concordance inequality is not verified.

@@

\textbf{When $n\geq 2$}, we get 
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + n}{a +n - 2} -1
	= \frac{n_0 - a + 2}{a +n - 2}
	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
\end{align}
so putting it all together we get 
\begin{align}
	\expect{\bregmanconj(\mu^*; \MAPm)} 
	& = \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
	&\leq \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
\end{align}

\textbf{When $n=1$} the bound~\eqref{eq:lemma_natp_bound}  on the expected natural parameter gives
\begin{align}
	\expect{\frac{\mu^*}{\MAPm}} - 1 
	\leq \frac{n_0 + 1}{a} -1
	= \frac{n_0 +1 - a}{a}
\end{align}
so putting it all together we get
\begin{align}
	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
\end{align}
so we do recover the same bias variance structure as when $n\geq2$.
\end{proof}
